String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-03 2023-1-19/sections/algorithms.tex
algorithm done

As mentioned in Section \ref{sec:overview}, the main idea of our approach is to model the counting operators symbolically by registers, instead of unfolding them explicitly. Additionally, we use registers to represent string lengths. After that, the satisfiability of string constraints involve regex-counting and string-length is reduced to the satisfiability of LIA, which can be solved by off-the-shelf SMT solvers. We also propose techniques to reduce automata sizes and utilize under approximations to enhance performance.
\subsection{Encode Regex with Counting to CEFA} \label{subsec:regex2cefa}
The most important point of our approach is encoding counting operators by registers rather than explicitly unfolding them. In this section, we discuss the encoding in detail. A counting operator is called \emph{handled} if it is not the sub-regex of complement, closure or other counting, and called \emph{unhandled} otherwise. For example, $(\Sigma \setminus a)^{\{m,n\}}$ is handled, while $((\Sigma \setminus a)^{\{m,n\}})^*$ is unhandled. We first discuss the encoding of handled counting operators, and then discuss the encoding of unhandled counting operators.

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without unhandled counting operator, the base case is constructing CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix , we only illustrate counting here.

Given a regex $\regex^{\{m,n\}}$ without unhandled counting operator, suppose the NFA recognizing $\regex$ is $\aut = (Q,\Sigma,\delta,q_I,F)$, then the CEFA recognizing $\regex^{\{m,n\}}$ is defined as $\aut_{m,n} = (Q, \Sigma, \delta',q_I, F\cup\{q_I\}, r, \theta)$ where:
\begin{itemize}
  \item $q_I$ is set to accepting state,
  \item $r$ is a new register,
  \item $\delta'$ is composed by transition $q\xrightarrow[(0)]{a} q'$ for each transition $q\xrightarrow[()]{a} q' \in \delta$, and transition $q_f\xrightarrow[(1)]{\epsilon} q_I$ for each accepting state $q_f\in F$,
  \item $\theta$ is the function mapping each accepting state to a same linear integer arithmetic $(m\leq )r\leq n$. The lower bound $m\leq r$ is unnecessary if $q_I\in F$.
\end{itemize}
The main idea is to add an epsilon transition to initial state from each accepting state like closure, to repeat the sub-regex. Then the counting time is incremented by 1 when the epsilon transition is invoked. Details are illustrated in Fig.\ref{fig:construct_repetition}. A new register $r$ is added to store the counting times. The transition $q\xrightarrow[(1)]{\epsilon} q_I$ it constructed to increment the counting times. The accepting condition $(m\leq) r\leq n$ is attached to each accepting state to constrain the counting times of the accepting word. After the construction, we need to eliminate epsilon transitions. It is done by adding transition $q_f\xrightarrow[(1)]{a} q$ for transition $q_I\xrightarrow[(0)]{a} q$ and then deleting the transition $q_f\xrightarrow[(1)]{\epsilon} q_I$, for each accepting state $q_f$.

@@

      \node at (-0.5,0.5) {$\scriptstyle (\myemph{m}\leq) r\leq \myemph{n}$};
      \node[state,initial,accepting]            (q_0)                      {};
      \node[state, accepting]         (q_1) [right=of q_0]       {};
      \path[->] (q_1) edge[bend right]   node    [above] {\myemph{$\scriptstyle \mathbf{\epsilon:(1)}$}} (q_0);
      \node [fit=(q_0) (q_1)] {$\cdots$};
    \end{tikzpicture}
    \caption{The CEFA recognizing $\regex^{\{m,n\}}$}
  \end{subfigure}
  \caption{The construction of counting operator}
  \label{fig:construct_repetition}
\end{figure}
\begin{example}
  To construct the CEFA recognizing $(\Sigma\setminus a)\{5,60\}$ with alphabet $\Sigma = \{a,b,c\}$. As shown in Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}, the base cases are the constructions of character $b$ (Fig.\ref{subfig:count_aut_b}) and $c$ (Fig.\ref{subfig:count_aut_c}). The first inductive step is construction the union of them (Fig.\ref{sub@subfig:count_aut_sigma_minus_a}). The second inductive step is to add counting information on the automaton. To do that we add a new register $r$ to save counting times,  transitions $q_2\xrightarrow[(1)]{\epsilon} q_0$ and $q_1\xrightarrow[(1)]{\epsilon} q_0$  to update the counting times. The accepting condition of each accepting state is $5\leq r\leq 60$, which ensure the counting time of the accepting word is in the range $[5,60]$. After epsilon elimination we get the CEFA recognizing $(\Sigma\setminus a)\{5,60\}$ (shown in Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}).
  \begin{figure}[h]
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]
        \node[state,initial]            (q_0)                      {$\scriptstyle q_0$};
        \node[state,accepting]                    (q_1) [right=of q_0]       {$\scriptstyle q_1$};

@@

Above discusses the construction of handled counting operators. The unhandled counting operator can not be constructed by the same way. For example, for the regex $(ab^{\{1,100\}})*$ with unhandled counting operator, there are infinite repetitions of $ab^{\{1,100\}}$. The counting times of each repetition are irrelevant, so that we can not represent them using finite registers. We have to syntactically rewrite the unhandled counting operator into union operator and concatenation operator. More exactly, the sub-regex $\regex^{\{m,n\}}$ in the unhandled regex is rewritten into $\underbrace{\regex\cdot\regex\cdots\regex}_{m \text{ times}}\mid\cdots\mid \underbrace{\regex\cdot\regex\cdots\regex}_{n \text{ times}}$.
% Based on the decision procedure defined in the paper \cite{atva2020}, we propose an efficient algorithm to solve ESL formula $\varphi\wedge \psi$ where $\varphi$ is the conjunction of literals and $\psi$ is the conjunction of linear literals without length operation. For each string variable $x$, we first compute pre-images of all linear literal $i=|x|$. Then we generate CEFAs of all regular expressions $\regex$ in all regular literals $x\in \regex$ and intersect these CEFAs and pre-images to get the final automaton $\aut_x$ for $x$. At present, there is a final CEFA for each string variable $x$. The linear arithmetic constraint $\psi$ restricts the accepting word of final CEFAs. Thus the satisfiability problem is the emptiness checking problem of final CEFAs under $\psi$. If it is empty, the string constraints are unsatisfiable. Otherwise, the string constraints are satisfiable.
% The emptiness checking problem of CEFAs under linear integer arithmetic is theoretically pspace-complete\cite{atva2020}. To solve it efficiently for a practical example, we develop heuristic ways such as under-approximation (Section \ref{subsec:under-approx}) and symbolic-aware simplification (Section \ref{subsec:simplify}).
\subsection{Handling of String-length}
Given a string-length constraint, we build a CEFA with a register storing the word length. Namely, the CEFA is $(\{q_I\}, \Sigma, \{q_I\xrightarrow[(1)]{\Sigma} q_I\}, q_I, \{q_I\}, r, \theta)$ where accpeting condition $\theta$ is based on the length formula. As an illustration, example \ref{eg:pre_len} shows the CEFA of length constraints $|x| > 120$, whose language is the set of all words with length greater than $120$. The simple loop $q_I\xrightarrow[(1)]{\Sigma} q_I$ increase $1$ on register $r$ for all char in $\Sigma$, so that the value of $r$ is equal to the length of word. The accepting condition $\theta$ attachs $q_I$ to $r > 120$, ensuring the length of accepting word is greater than $120$.

@@

      \path[->] (q_0) edge [loop below] node{$\scriptstyle\Sigma:(1)$} ();
    \end{tikzpicture}
  \end{figure}
\end{example} 
% \subsection{Combine Regex-counting and String-length}
% In the previous section, we have discussed the construction of regex-counting operator and string-length constraint. In this section, we show how to combine them to deal with the string constraints with regex-counting and string-length. 
% \begin{definition} [The Product of Two CEFA]

@@

As mentioned in \cite{atva2020}, the \emptinessprob{} is theoretically Pspace-complete. We has used a model-checking tool \emph{nuXmv}\cite{nuxmv} to solve it in previous work. However, it is not efficient enough for complex string constraints. So we put forward a new framework that brings in symbolic-aware simplification and under-approximation heuristics.

The main idea of our framework is to simplify the input CEFAs firstly and under-approximate the solution of the \emptinessprob{} by adding length bounds on string variables. If the under-approximation is satisfiable, the CEFAs are not empty under $\psi$. Otherwise, we check the Parikh image of the simplified CEFAs. If the Parikh image is satisfiable, the CEFAs are not empty under $\psi$. Or else, they are empty. The framework is \emph{sound} and \emph{complete}.

The framework is shown in Algorithm \ref{alg:emptiness}, we simplify the input CEFAs at line 1 because they may have too many transitions and registers to solve. The main idea of the simplification is treating the alphabet as unary and the symbolic updates as characters of transitions in the minimization. After the simplification, we try to find a solution by under-approximation at line 2 and line 3. The under-approximation algorithm explore possible satisfiable register values within a length bound $MaxBound$. If it finds a solution then we return true because the under-approximation is sound for sat. Otherwise, we compute the Parikh images of the simplified CEFAs and check if the Parikh images are satisfiable in conjunction with $\psi$ at line 12. The Parikh image checking is complete so that satisfiability implies non-emptiness directly.

The under-approximation program is incremental. In other words, it tries to find a solution with smaller length bound of each string variable first and increment the string length bound if the search of smaller length bound fails. $MaxBound$ is the max length bound we try which is empirically set to $15$ in our experiment. It should not be too large because of the exponential growth of searching space. The details of simplification and under-approximation are shown in Algorithm \ref{alg:simplify} and the details of  are shown in Algorithm \ref{alg:underApprox}.

\begin{algorithm}[h]
  \caption{ $\algfun{nonEmpty}(auts, \psi)$}
  \label{alg:emptiness}
  \begin{algorithmic}[1]
    \Require the CEFAs $auts$ and the linear integer arithmetic $\psi$
    \Ensure $true$ or $false$
    \Statex
    \State $simpliAuts \leftarrow \algfun{simplify}(auts)$
    % \For{$bound \gets 1, MaxBound$ }
    % \State $\varphi_{under}\gets \algfun{underApprox}(simpliAuts, bound)$
    % \If{$\varphi_{under}\wedge \psi$ is sat}
    % \State \textbf{return} $true$
    % \EndIf
    % \EndFor
    \If{\algfun{underApproxSolve}($simpliAuts, \psi, MaxBound$) is true}
    \State \textbf{return} $true$
    \EndIf
    \State $\varphi \gets \algfun{parikhImage}(simpliAuts)$
    \If{$\varphi\wedge \psi$ is sat}
    \State \textbf{return} $true$
    \Else
    \State \textbf{return} $false$
    \EndIf
  \end{algorithmic}
\end{algorithm}
\subsubsection{Symbolic-Aware Simplification} \label{subsec:simplify}
The purpose of simplification is to remove duplicated transitions and registers. In the emptiness checking Algorithm \ref{alg:emptiness}, the non-emptiness is equal to the reachability of CEFAs. Thus, the symbolic updates on transitions are meritorious because it influences registers values, while the characters are not. The alphabet of the CEFA can be seen as unary (i.e., the alphabet is $\{a\}$) and the vectors as characters. The special vector $\myvec{0}_n$ with only zero elements is seen as $\epsilon$ character. Namely, given a CEFA $\aut = (Q, \Sigma, \delta, q_I, F, R, \theta)$, we obtain a symbolic form  $\aut_{sym} = (Q, \Sigma', \delta', q_I, F, R, \theta)$ where
\begin{itemize}
  \item $\Sigma' = \{a\}$,
  \item the transition set $\delta'$ is composed of transition $q\xrightarrow[\myvec{v}]{a}q'$ if there is a transition $q\xrightarrow[\myvec{v}]{b} q'$ in $\delta$ for $b\in \Sigma$.
\end{itemize}

 For the symbolic form $\aut_{sym}$, we treat $q\xrightarrow[\myvec{0}_n]{a}q'$ as epsilon transition and compute the epsilon-closure based on it. Likewise, we treat $q\xrightarrow[\myvec{v}]{a}q'$ as a transition with character $(a, \myvec{v})$ and determine the automaton based on it. Two accepting states are equivalent if they have the same accepting condition, and two states are equivalent if they reach equivalent states for each character. We simplify the automaton by merging these equivalent states. Briefly, we directly apply algorithms of epsilon-closure, determination, and simplification in \cite{aut_hopcraft}.
The explanation above discuss funcitons $\algfun{determinizeByVec}$ and $\algfun{minimizeByVec}$ at line 3 and line 4. From line 5 to line 13, we check whether two registers $r_i$ and $r_j$ are duplicated. (i.e., the symbolic updates of them are always the same). If them are duplicated at line 6, we remove one of them at line 7 and the corresponding symbolic update at line 9. After deleting the duplicated registers, we add constraint $r_i = r_j$ to store the value of the deleted register, to ensure the value of it is consistent at line 11.

We have the following proposition:
\begin{proposition}
  Given a set of CEFAs $\Lambda=\{\aut_1, \cdots, \aut_2\}$ and arbitrary LIA $\psi$, the emptiness of $\Lambda$ on $\psi$ is equal to the emptiness of $\algfun{simplify}(\Lambda)$ on $\psi$.
\end{proposition}
\begin{algorithm}[h]
  \caption{$\algfun{simplify}(auts)$}
  \label{alg:simplify}
  \begin{algorithmic}[1]
    \Require A set of CEFAs
    \Ensure The simplified CEFAs
    \Statex
    \For{$\aut \in auts$}
    \State Suppose that $\aut = (Q, \Sigma, \delta, q_I, F,
      R, \theta)$ and $R=(r_1,\cdots, r_n)$
    \State $\algfun{determinizeByVec}(\aut)$
    \State $\algfun{minimizeByVec}(\aut)$
    \ForAll{$(i, j)$ where $1\leq i \leq j \leq n$}
    \If {$\myvec{v}[i] = \myvec{v}[j]$ for every vector $\myvec{v}\in \aut$ }
    \State $R\gets(\cdots r_{j-1}r_{j+1}\cdots)$ \Comment remove $r_j$ from the register vector $R$
    \ForAll{$\myvec{v'}\in \aut$}
    \State $\myvec{v'}\gets(\cdots\myvec{v'}[j-1]\myvec{v'}[j+1]\cdots)$ \Comment remove $\myvec{v'}[j]$ from each vector $\myvec{v'}$
    \EndFor
    \State $\theta \gets\theta\wedge r_i = r_j$
    \EndIf
    \EndFor
    \EndFor
    \State \textbf{return} $auts$
  \end{algorithmic}
\end{algorithm}
\begin{example}
  Considering the CEFA (Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}) recognizing $(\Sigma\setminus a)\{5,60\}$ where $\Sigma = \{a,b,c\}$. We first obtain a symbolic form of it by using unary alphabet (Fig.\ref{subfig:sigma_minus_a_5_60:simplify}) and then minimize the symbolic form based on the label $"a:(1)"$(Fig.\ref{subfig:simp_final}).
  \begin{figure}[h]
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=1cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.005cm},
        ]
        \node at (-1, 1) {$\scriptstyle 5\leq r \leq 60$};
        \node[state,initial, accepting]      (q_0)                      {$\scriptstyle q_0$};
        \node (q_tmp) [right=of q_0]       {};
        \node[state, accepting]   (q_1) [above=of q_tmp]       {$\scriptstyle q_1$};
        \node[state, accepting]   (q_2) [below=of q_tmp]       {$\scriptstyle q_2$};

        \path[->] (q_0) edge node [right] {$\scriptstyle a:(0)$} (q_1)
        (q_1) edge [loop right] node {$\scriptstyle a:(1)$} (q_1)
        (q_0) edge           node [right] {$\scriptstyle a:(0)$} (q_2)
        (q_2) edge [loop right] node {$\scriptstyle a:(1)$} (q_0);
      \end{tikzpicture}
      \caption{The symbolic form of $\aut_{\Sigma\setminus a\{5,60\}}$}
      \label{subfig:sigma_minus_a_5_60:simplify}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=1cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]

        \node at (-0.5, 0.5) {$\scriptstyle 5\leq r \leq 60$};
        \node[state,initial,accepting]            (q_0)       {$\scriptstyle q_I$};

        \path[->] (q_0) edge [loop below] node{$\scriptstyle a:(1)$} ();
      \end{tikzpicture}
      \caption{Simplify the CEFA based on symbolic label}
      \label{subfig:simp_final}
    \end{subfigure}
    \caption{Symbolic simplification removes duplicated transitions and states.}
    \label{fig:simplification_example}
  \end{figure}

>>>>

As mentioned in Section \ref{sec:overview}, the main idea of our approach is to model the counting operators symbolically by registers instead of unfolding them explicitly. Additionally, we use registers to represent string lengths. After that, the satisfiability of string constraints involves regex-counting, and string-length is reduced to the satisfiability of LIA, which off-the-shelf SMT solvers can solve. We also propose techniques to reduce automata sizes and utilize under approximations to enhance performance.
\subsection{Encode Regex with Counting to CEFA} \label{subsec:regex2cefa}
The most important point of our approach is encoding counting operators by registers rather than explicitly unfolding them. In this section, we discuss the encoding in detail. A counting operator is called \emph{handled} if it is not the sub-regex of complement, closure, or another counting, and called \emph{unhandled} otherwise. For example, $(\Sigma \setminus a)^{\{m,n\}}$ is handled, while $((\Sigma \setminus a)^{\{m,n\}})^*$ is unhandled. We first discuss the encoding of handled counting operators and then the encoding of unhandled counting operators.

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without an unhandled counting operator, the base case is constructing the CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement, and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix , we only illustrate counting here.

Given a regex $\regex^{\{m,n\}}$ without unhandled counting operator, suppose the NFA recognizing $\regex$ is $\aut = (Q,\Sigma,\delta,q_I,F)$, then the CEFA recognizing $\regex^{\{m,n\}}$ is defined as $\aut_{m,n} = (Q, \Sigma, \delta',q_I, F\cup\{q_I\}, r, \theta)$ where:
\begin{itemize}
  \item $q_I$ is set to accepting state,
  \item $r$ is a new register,
  \item $\delta'$ is composed by transition $q\xrightarrow[(0)]{a} q'$ for each transition $q\xrightarrow[()]{a} q' \in \delta$, and transition $q_f\xrightarrow[(1)]{\epsilon} q_I$ for each accepting state $q_f\in F$,
  \item $\theta$ is the function mapping each accepting state to the same linear integer arithmetic $(m\leq )r\leq n$. The lower bound $m\leq r$ is unnecessary if $q_I\in F$.
\end{itemize}
The main idea is to add an epsilon transition to the initial state from each accepting state, like closure, to repeat the sub-regex. Then the counting time is incremented by 1 when the epsilon transition is invoked. Details are illustrated in Fig.\ref{fig:construct_repetition}. A new register $r$ is added to store the counting times. The transition $q\xrightarrow[(1)]{\epsilon} q_I$ is constructed to increment the counting times. The accepting condition $(m\leq) r\leq n$ is attached to each accepting state to constrain the counting times of the accepting word. After the construction, we need to eliminate epsilon transitions. It is done by adding transition $q_f\xrightarrow[(1)]{a} q$ for transition $q_I\xrightarrow[(0)]{a} q$ and then deleting the transition $q_f\xrightarrow[(1)]{\epsilon} q_I$, for each accepting state $q_f$.

@@

      \node at (-0.5,0.5) {$\scriptstyle (\myemph{m}\leq) r\leq \myemph{n}$};
      \node[state,initial,accepting]            (q_0)                      {};
      \node[state, accepting]         (q_1) [right=of q_0]       {};
      \path[->] (q_1) edge[bend right]   node    [above] {\myemph{$\scriptstyle \mathbf{\epsilon:(1)}$}} (q_0);
      \node [fit=(q_0) (q_1)] {$\cdots$};
    \end{tikzpicture}
    \caption{The CEFA recognizing $\regex^{\{m,n\}}$}
  \end{subfigure}
  \caption{The construction of counting operator}
  \label{fig:construct_repetition}
\end{figure}
\begin{example}
  To construct the CEFA recognizing $(\Sigma\setminus a)\{5,60\}$ with alphabet $\Sigma = \{a,b,c\}$. As shown in Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}, the base cases are the constructions of character $b$ (Fig.\ref{subfig:count_aut_b}) and $c$ (Fig.\ref{subfig:count_aut_c}). The first inductive step is building their union (Fig.\ref{sub@subfig:count_aut_sigma_minus_a}). The second inductive step is to add counting information on the automaton. To do that we add a new register $r$ to save counting times,  transitions $q_2\xrightarrow[(1)]{\epsilon} q_0$ and $q_1\xrightarrow[(1)]{\epsilon} q_0$  to update the counting times. The accepting condition of each accepting state is $5\leq r\leq 60$, which ensures the accepting word's counting time is in the range $[5,60]$. After epsilon elimination we get the CEFA recognizing $(\Sigma\setminus a)\{5,60\}$ (shown in Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}).
  \begin{figure}[h]
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=2cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]
        \node[state,initial]            (q_0)                      {$\scriptstyle q_0$};
        \node[state,accepting]                    (q_1) [right=of q_0]       {$\scriptstyle q_1$};

@@

Above discusses the construction of handled counting operators. In the same way, we can not construct the unhandled counting operator. For example, for the regex $(ab^{\{1,100\}})*$ with unhandled counting operator, there are infinite repetitions of $ab^{\{1,100\}}$. The counting times of each repetition are irrelevant, so we can not represent them using finite registers. We must syntactically rewrite the unhandled counting operator into union and concatenation operators. More exactly, the sub-regex $\regex^{\{m,n\}}$ in the unhandled regex is rewritten into $\underbrace{\regex\cdot\regex\cdots\regex}_{m \text{ times}}\mid\cdots\mid \underbrace{\regex\cdot\regex\cdots\regex}_{n \text{ times}}$.
% Based on the decision procedure defined in the paper \cite{atva2020}, we propose an efficient algorithm to solve ESL formula $\varphi\wedge \psi$ where $\varphi$ is the conjunction of literals and $\psi$ is the conjunction of linear literals without length operation. For each string variable $x$, we first compute pre-images of all linear literal $i=|x|$. Then we generate CEFAs of all regular expressions $\regex$ in all regular literals $x\in \regex$ and intersect these CEFAs and pre-images to get the final automaton $\aut_x$ for $x$. There is a final CEFA for each string variable $x$. The linear arithmetic constraint $\psi$ restricts the accepting word of final CEFAs. Thus the satisfiability problem is the emptiness checking problem of final CEFAs under $\psi$. If it is empty, the string constraints are unsatisfiable. Otherwise, the string constraints are satisfiable.
% The emptiness checking problem of CEFAs under linear integer arithmetic is theoretically pspace-complete\cite{atva2020}. To solve it efficiently for a practical example, we develop heuristic ways such as under-approximation (Section \ref{subsec:under-approx}) and symbolic-aware simplification (Section \ref{subsec:simplify}).
\subsection{Handling of String-length}
Given a string-length constraint, we build a CEFA with a register storing the word length. Namely, the CEFA is $(\{q_I\}, \Sigma, \{q_I\xrightarrow[(1)]{\Sigma} q_I\}, q_I, \{q_I\}, r, \theta)$ where accpeting condition $\theta$ is based on the length formula. As an illustration, example \ref{eg:pre_len} shows the CEFA of length constraints $|x| > 120$, whose language is the set of all words with a length greater than $120$. The simple loop $q_I\xrightarrow[(1)]{\Sigma} q_I$ increase $1$ on register $r$ for all char in $\Sigma$, so that the value of $r$ is equal to the length of the accepting word. The accepting condition $\theta$ attaches $q_I$ to $r > 120$, ensuring the length of the accepting word is greater than $120$.

@@

      \path[->] (q_0) edge [loop below] node{$\scriptstyle\Sigma:(1)$} ();
    \end{tikzpicture}
  \end{figure}
\end{example} 
% \subsection{Combine Regex-counting and String-length}
% In the previous section, we discussed the construction of the regex-counting operator and string-length constraint. In this section, we show how to combine them to deal with the string constraints with regex-counting and string-length. 
% \begin{definition} [The Product of Two CEFA]

@@

As mentioned in \cite{atva2020}, the \emptinessprob{} is theoretically Pspace-complete. In previous work, we have used a model-checking tool \emph{nuXmv}\cite{nuxmv} to solve it. However, it needs to be more efficient for complex string constraints. So we put forward a new framework that brings in symbolic-aware simplification and under-approximation heuristics.

The main idea of our framework is to simplify the input CEFAs firstly and under-approximate the solution of the \emptinessprob{} by adding length bounds on string variables. If the under-approximation is satisfiable, the CEFAs are not empty under $\psi$. Otherwise, we check the Parikh image of the simplified CEFAs. If the Parikh image is satisfiable, the CEFAs are not empty under $\psi$. Alternatively, they are empty. The framework is \emph{sound} and \emph{complete}.

The framework is shown in Algorithm \ref{alg:emptiness}. We simplify the input CEFAs at line 1 because they may have too many transitions and registers to solve. The main idea of the simplification is treating the alphabet as unary and the symbolic updates as characters of transitions in the minimization. After the simplification, we try to find a solution by under-approximation at line 2 and line 3. The under-approximation algorithm explores possible satisfiable register values within a length bound $MaxBound$. We return true if it finds a solution because the under-approximation is sound for sat. Otherwise, we compute the Parikh images of the simplified CEFAs and check if the Parikh images are satisfiable in conjunction with $\psi$ at line 12. The Parikh image checking is complete so that satisfiability implies non-emptiness directly.

The under-approximation program is incremental. In other words, it first tries to find a solution with a smaller length bound of each string variable and increments the string length bound if the search for a smaller length bound fails. $MaxBound$ is the max length bound we try, empirically set to $15$ in our experiment. It should not be too large because of the exponential growth of searching space. The details of simplification and under-approximation are shown in Algorithm \ref{alg:simplify}, and the details are shown in Algorithm \ref{alg:underApprox}.

\begin{algorithm}[h]
  \caption{ $\algfun{nonEmpty}(auts, \psi)$}
  \label{alg:emptiness}
  \begin{algorithmic}[1]
    \Require the CEFAs $auts$ and the linear integer arithmetic $\psi$
    \Ensure $true$ or $false$
    \Statex
    \State $simpliAuts \leftarrow \algfun{simplify}(auts)$
    % \For{$bound \gets 1, MaxBound$ }
    % \State $\varphi_{under}\gets \algfun{underApprox}(simpliAuts, bound)$
    % \If{$\varphi_{under}\wedge \psi$ is sat}
    % \State \textbf{return} $true$
    % \EndIf
    % \EndFor
    \If{\algfun{underApproxSolve}($simpliAuts, \psi, MaxBound$) is true}
    \State \textbf{return} $true$
    \EndIf
    \State $\varphi \gets \algfun{parikhImage}(simpliAuts)$
    \If{$\varphi\wedge \psi$ is sat}
    \State \textbf{return} $true$
    \Else
    \State \textbf{return} $false$
    \EndIf
  \end{algorithmic}
\end{algorithm}
\subsubsection{Symbolic-Aware Simplification} \label{subsec:simplify}
The purpose of simplification is to remove duplicated transitions and registers. In the emptiness checking Algorithm \ref{alg:emptiness}, the non-emptiness equals the reachability of CEFAs. The symbolic updates on transitions are meritorious because it influences register values, while the characters are not. The alphabet of the CEFA can be ignored (i.e., the alphabet is $\{a\}$), and the vectors can be seen as characters. The special vector $\myvec{0}_n$ with only zero elements is seen as a $\epsilon$ character since it does not change the values of registers. Namely, given a CEFA $\aut = (Q, \Sigma, \delta, q_I, F, R, \theta)$, we obtain a symbolic NFA  $\aut_{sym} = (Q, \Sigma', \delta', q_I, F)$ where
\begin{itemize}
  \item $\Sigma' = \{\myvec{v} \mid \text{there exists one transition in } \aut \text{ with update } \myvec{v}\}$,
  \item the transition set $\delta'$ is composed of transition $q\xrightarrow{\myvec{v}}q'$ if there is a transition $q\xrightarrow[\myvec{v}]{b} q'$ in $\delta$.
\end{itemize}

 For the symbolic NFA $\aut_{sym}$, we treat $q\xrightarrow{\myvec{0}_n} q'$ as an epsilon transition since it does not change the values of the registers. Then we apply the epsilon elimination algorithm \cite{aut_hopcraft} to delete these epsilon transitions. After that, we apply the determination and minimization algorithm \cite{aut_hopcraft} of NFAs to $\aut_{sym}$, resulting in a smaller automaton $\aut_{simp}$. Finally, we reinterpret $\aut_{simp}$ as a CEFA with the accepting condition of $\aut$ and unary alphabet $\Sigma = \{a\}$. Above illustrates functions $\algfun{determinizeByVec}$ and $\algfun{minimizeByVec}$ from line 3 to line 5. From line 6 to line 14, we check whether two registers $r_i$ and $r_j$ are duplicated. (i.e., the symbolic updates of them are always the same). If they are duplicated at line 6, we remove one of them and the corresponding symbolic update from line 8 to line 10. After deleting the duplicated registers, we add constraint $r_i = r_j$ to store the value of the deleted register at line 12, ensuring that the value of the deleted register is consistent.

We have the following proposition:
\begin{proposition}
  Given a set of CEFAs $\Lambda=\{\aut_1, \cdots, \aut_2\}$ and arbitrary LIA $\psi$, the emptiness of $\Lambda$ on $\psi$ is equal to the emptiness of $\algfun{simplify}(\Lambda)$ on $\psi$.
\end{proposition}
\begin{algorithm}[h]
  \caption{$\algfun{simplify}(auts)$}
  \label{alg:simplify}
  \begin{algorithmic}[1]
    \Require A set of CEFAs
    \Ensure The simplified CEFAs
    \Statex
    \For{$\aut \in auts$}
    \State Suppose that $\aut = (Q, \Sigma, \delta, q_I, F,
      R, \theta)$ and $R=(r_1,\cdots, r_n)$
    \State $\algfun{epsilonElimByVec}(\aut)$
    \State $\algfun{determinizeByVec}(\aut)$
    \State $\algfun{minimizeByVec}(\aut)$
    \ForAll{$(i, j)$ where $1\leq i \leq j \leq n$}
    \If {$\myvec{v}[i] = \myvec{v}[j]$ for every vector $\myvec{v}\in \aut$ }
    \State $R\gets(\cdots r_{j-1}r_{j+1}\cdots)$ \Comment remove $r_j$ from the register vector $R$
    \ForAll{$\myvec{v'}\in \aut$}
    \State $\myvec{v'}\gets(\cdots\myvec{v'}[j-1]\myvec{v'}[j+1]\cdots)$ \Comment remove $\myvec{v'}[j]$ from each vector $\myvec{v'}$
    \EndFor
    \State $\theta \gets\theta\wedge r_i = r_j$
    \EndIf
    \EndFor
    \EndFor
    \State \textbf{return} $auts$
  \end{algorithmic}
\end{algorithm}
\begin{example}
  Considering the CEFA (Fig.\ref{subfig:count_aut_sigma_minus_a_5_60}) recognizing $(\Sigma\setminus a)\{1,60\}$ where $\Sigma = \{a,b,c\}$. We first obtain a symbolic NFA of it by ignoring the unary alphabet (Fig.\ref{subfig:sigma_minus_a_5_60:simplify}). Then we eliminate epsilon transitions $q_0\xrightarrow[]{(0)} q_1$ and $q_0\xrightarrow[]{(0)}q_2$, resulting in the minimized NFA directly. Finally, we reinterpret it as CEFA by adding accepting condition $1\leq r \leq 60$(Fig.\ref{subfig:simp_final}).
  \begin{figure}[h]
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=1cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.005cm},
        ]
        % \node at (-1, 1) {$\scriptstyle 5\leq r \leq 60$};
        \node[state,initial, accepting]      (q_0)                      {$\scriptstyle q_0$};
        \node (q_tmp) [right=of q_0]       {};
        \node[state, accepting]   (q_1) [above=of q_tmp]       {$\scriptstyle q_1$};
        \node[state, accepting]   (q_2) [below=of q_tmp]       {$\scriptstyle q_2$};

        \path[->] (q_0) edge node [right] {$(0)$} (q_1)
        (q_1) edge [loop right] node {$(1)$} (q_1)
        (q_0) edge           node [right] {$(0)$} (q_2)
        (q_2) edge [loop right] node {$(1)$} (q_0);
      \end{tikzpicture}
      \caption{The symbolic NFA of $\aut_{\Sigma\setminus a\{5,60\}}$}
      \label{subfig:sigma_minus_a_5_60:simplify}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \begin{tikzpicture}[
        shorten >=1pt,node distance=1cm,on grid,>={Stealth[round]},
        initial text=, every state/.style={minimum size = 0.001cm},
        ]

        \node at (-0.5, 0.5) {$\scriptstyle 1\leq r \leq 60$};
        \node[state,initial,accepting]            (q_0)       {$\scriptstyle q_I$};

        \path[->] (q_0) edge [loop below] node{$(1)$} ();
      \end{tikzpicture}
      \caption{Simplify the CEFA based on symbolic label}
      \label{subfig:simp_final}
    \end{subfigure}
    \caption{Symbolic simplification removes duplicated transitions and states.}
    \label{fig:simplification_example}
  \end{figure}
